Weighted Model Counting with Conditional Weights for Bayesian Networks

### 2021-02-09 doc/paper2/paper.tex
Tables, figures, etc.

\section{Introduction}

TODO: introduce (also talk more about how Cachet and Ace are used)
\begin{itemize}
\item \texttt{bklm16} \citep{DBLP:conf/ecai/BartKLM16}
\item \textsf{query-dnnf}
  \footnote{\url{http://www.cril.univ-artois.fr/kc/d-DNNF-reasoner.html}}
\item \textsf{DPMC} \citep{DBLP:conf/cp/DudekPV20}
\item tackle a comparable problem, but their encoding was never implemented or
  empirically tested (i.e., the implementation is left as future work)
  \citep{DBLP:conf/scai/VomlelT13}
\end{itemize}

@@

\section{Related Work} \label{sec:related}

@@

\subsection{The Space of Functions on Boolean Algebras}

@@

Since many measures of interest may not be factorable, a well-known way to
encode them into instances of WMC is by adding more literals
\citep{DBLP:journals/ai/ChaviraD08}. We can use the measure-theoretic
perspective on WMC to show that this is always possible, however, as ensuing
sections will demonstrate, it often makes the inference task much harder in
practice.\footnote{The proofs of this and other theoretical results can be found
  in the technical appendix.}

@@

First, \textsf{ADDMC} constructs the \emph{Gaifman graph}
of the input CNF formula as an aid for the algorithm's
heuristics. This graph has as vertices the variables of the formula, and there
is an edge between two variables $u$ and $v$ if there is a clause in the formula
that contains both $u$ and $v$. We extend this definition to functions on
Boolean algebras, i.e., the factors of $\phi$. For any pair of distinct
variables $u, v \in U$, we draw an edge between them in the Gaifman graph if
there is a function $\alpha\colon 2^X \to \mathbb{R}_{\ge 0}$ that is a factor
of $\phi$ such that $u \in X$ and $v \in X$. For instance, a factor such as
$\mathrm{CPT}_X$ will enable edges between all distinct pairs of variables in
$\mathcal{E}^*(X)$. Second, even though the function $\phi$ produced by
\cref{alg:encoding} is constructed to have $2^U$ as its domain, sometimes the
domain is effectively reduced to $2^V$ for some $V \subset U$ by the ADD
manipulation algorithms that optimise the ADD representation of a function. For
a simple example, consider $\alpha: 2^{\{a\}} \to \mathbb{R}_{\ge 0}$ defined as
$\alpha(\{a\}) = \alpha(\emptyset) = 0.5$. Then $\alpha$ can be reduced to
$\alpha'\colon 2^{\emptyset} \to \mathbb{R}_{\ge 0}$ defined as
$\alpha'(\emptyset) = 0.5$. To compensate for these reductions, for the original
WMC format with a weight function $w\colon U \cup \{ \neg u \mid u \in U \} \to
\mathbb{R}_{\ge 0}$, \textsf{ADDMC} would multiply its computed answer by
$\prod_{u \in U \setminus V} w(u) + w(\neg u)$. With the new WMC format, we
instead multiply the answer by $2^{|U \setminus V|}$. Each `excluded' variable
$u \in U \setminus V$ satisfies two properties: all weights associated with $u$
are equal to $0.5$ (otherwise the corresponding CPT would depend on $u$, and $u$
would not be excluded), and all other CPTs are independent of $u$ (or they may
have a trivial dependence, where the probability stays the same if $u$ is
replaced with its complement). Thus, the CPT that corresponds to $u$ still
multiplies the weight of every atom in the Boolean algebra by $0.5$, but the
number of atoms under consideration is halved. To correct for this, we multiply
the final answer by two for every $u \in U \setminus V$.

\section{Experimental Comparison} \label{sec:experiments}

In this section, we describe an experimental study comparing the five WMC
encodings for Bayesian networks when run with \textsf{ADDMC}. The experiments
were run on servers with Intel Xeon Gold 6138 and Intel Xeon E5-2630
processors\footnote{Each instance is run on the same processor for all
  encodings.} running Scientific~Linux~7 with an \SI{8}{\gibi\byte} memory limit
and a \SI{1000}{\second} timeout.

@@

\begin{figure*}[t]
  \centering
  \input{cumulative.tex}%
  \caption{Cumulative number of instances solved by \textsf{ADDMC} over time
    using each encoding.}
  \label{fig:cumulative}
\end{figure*}

\begin{table} % TODO: update or remove
  \centering
  \caption{The numbers of instances (out of 1466) solved by \textsf{ADDMC} with
    each encoding (uniquely, faster than with others, and in total) under the
    described constraints.}%
  \begin{tabular}{lrrr}
    \toprule
    Encoding & Unique & Fastest & Total \\
    \midrule
    \texttt{cd05} & 2 & 3 & 380 \\
    \texttt{cd06} & 0 & 0 & 363 \\
    \texttt{cw} & 234 & 935 & 980 \\
    \texttt{d02} & 40 & 111 & 797 \\
    \texttt{sbk05} & 12 & 44 & 765 \\
    \bottomrule
  \end{tabular}
  \label{tbl:tallies}
\end{table}

\begin{figure*}[t]
  \centering
  \input{scatter.tex}%
  \caption{\textsf{ADDMC} inference time using \texttt{cw} compared to
    \texttt{d02} (left) and \texttt{sbk05} (right) for each instance in all data
    sets.}
  \label{fig:scatter}
\end{figure*}

@@

\begin{table}
  \centering
  \caption{Asymptotic upper bounds on the numbers of variables and clauses/ADDs
    for each encoding.}
  \label{tbl:asymptotes}
  \begin{tabular}{lcc}
    \toprule
    Encoding(s) & Variables & Clauses/ADDs \\
    \midrule
    \texttt{cd05}, \texttt{cd06}, \texttt{sbk05} & $O(nv^{d+1})$ & $O(nv^{d+1})$ \\
    \texttt{cw} & $O(nv)$ & $O(nv^2)$ \\
    \texttt{d02} & $O(nv^{d+1})$ & $O(ndv^{d+1})$ \\
    \bottomrule
  \end{tabular}
\end{table}

@@

\clearpage
\bibliography{paper}

>>>>

\section{Introduction}
% TODO: introduce \textsf{DPMC} \citep{DBLP:conf/cp/DudekPV20}

@@

\section{Related Work} \label{sec:related}
% TODO: tackle a comparable problem, but their encoding was never implemented or
% empirically tested (i.e., the implementation is left as future work)
% \citep{DBLP:conf/scai/VomlelT13}

% TODO: introduce \texttt{bklm16} \citep{DBLP:conf/ecai/BartKLM16}

@@

\subsection{The Space of Functions on Boolean Algebras}
% TODO: properly define what logical operations on ADDs do. Mistake in the
% paper: an ADD is just a pseudo-boolean function: All operations on ADDs
% (whatever they are) are pointwise (perhaps /\, \/, ~ may be exceptions).

@@

Since many measures of interest may not be factorable, a well-known way to
encode them into instances of WMC is by adding more literals
\citep{DBLP:journals/ai/ChaviraD08}. We can use the measure-theoretic
perspective on WMC to show that this is always possible, however, as ensuing
sections will demonstrate, it often makes the inference task much harder in
practice.\footnote{The proofs of this and other theoretical results can be found
  in the appendix.}

@@

First, \textsf{ADDMC} constructs the \emph{primal} (a.k.a. Gaifman) \emph{graph}
of the input CNF formula as an aid for the algorithm's heuristics. This graph
has as vertices the variables of the formula, and there is an edge between two
variables $u$ and $v$ if there is a clause in the formula that contains both $u$
and $v$. We extend this definition to functions on Boolean algebras, i.e., the
factors of $\phi$. For any pair of distinct variables $u, v \in U$, we draw an
edge between them in the primal graph if there is a function $\alpha\colon 2^X
\to \mathbb{R}_{\ge 0}$ that is a factor of $\phi$ such that $u \in X$ and $v
\in X$. For instance, a factor such as $\mathrm{CPT}_X$ will enable edges
between all distinct pairs of variables in $\mathcal{E}^*(X)$. Second, even
though the function $\phi$ produced by \cref{alg:encoding} is constructed to
have $2^U$ as its domain, sometimes the domain is effectively reduced to $2^V$
for some $V \subset U$ by the ADD manipulation algorithms that optimise the ADD
representation of a function. For a simple example, consider $\alpha: 2^{\{a\}}
\to \mathbb{R}_{\ge 0}$ defined as $\alpha(\{a\}) = \alpha(\emptyset) = 0.5$.
Then $\alpha$ can be reduced to $\alpha'\colon 2^{\emptyset} \to \mathbb{R}_{\ge
  0}$ defined as $\alpha'(\emptyset) = 0.5$. To compensate for these reductions,
for the original WMC format with a weight function $w\colon U \cup \{ \neg u
\mid u \in U \} \to \mathbb{R}_{\ge 0}$, \textsf{ADDMC} would multiply its
computed answer by $\prod_{u \in U \setminus V} w(u) + w(\neg u)$. With the new
WMC format, we instead multiply the answer by $2^{|U \setminus V|}$. Each
`excluded' variable $u \in U \setminus V$ satisfies two properties: all weights
associated with $u$ are equal to $0.5$ (otherwise the corresponding CPT would
depend on $u$, and $u$ would not be excluded), and all other CPTs are
independent of $u$ (or they may have a trivial dependence, where the probability
stays the same if $u$ is replaced with its complement). Thus, the CPT that
corresponds to $u$ still multiplies the weight of every atom in the Boolean
algebra by $0.5$, but the number of atoms under consideration is halved. To
correct for this, we multiply the final answer by two for every $u \in U
\setminus V$.

\section{Experimental Comparison} \label{sec:experiments}
% TODO: introduce \textsf{query-dnnf}
% \footnote{\url{http://www.cril.univ-artois.fr/kc/d-DNNF-reasoner.html}} (also
% talk more about how Cachet and Ace are used)
% TODO: included original WMC algorithms
% TODO: Even instances that are solved in less than 0.1 s are too big for edge
% count to be efficiently calculated.
% TODO: my plot's sbk05 data should match (and matches) the ADDMC/DPMC vs c2d
% comparison in the paper (although my sbk05 data only roughly corresponds to
% half of their dataset)
% TODO: mention that both encoding and inference times are limited to 1000 s
% TODO: mention that we don't run ADDMC+cd05 and ADDMC+cd06 because they don't produce correct answers

\begin{figure*}[t]
  \centering
  \input{cumulative.tex}%
  \caption{Cumulative numbers of instances solved by combinations of algorithms
    and encodings over time.}
  \label{fig:cumulative}
\end{figure*}

\begin{table}
  \centering
  \caption{The numbers of instances (out of 1466) solved by
    each combination of algorithm and encoding (uniquely, faster than others, and in total).}%
  \begin{tabular}{lrrr}
    \toprule
    Algorithm \& Encoding & Unique & Fastest & Total \\
    \midrule
    $\textsf{Ace} + \texttt{cd05}$ & \textcolor{gray}{0} & 55 & 1169 \\
    $\textsf{Ace} + \texttt{cd06}$ & \textcolor{red}{34} & 218 & \textcolor{red}{1259} \\
    $\textsf{Ace} + \texttt{d02}$ & \textcolor{gray}{0} & 46 & 993 \\
    $\textsf{ADDMC} + \texttt{bklm16}$ & \textcolor{gray}{0} & 29 & \textcolor{blue}{617} \\
    $\textsf{ADDMC} + \texttt{cw}$ & 14 & \textcolor{red}{770} & 919 \\
    $\textsf{ADDMC} + \texttt{d02}$ & \textcolor{gray}{0} & \textcolor{gray}{0} & 703 \\
    $\textsf{ADDMC} + \texttt{sbk05}$ & \textcolor{gray}{0} & \textcolor{gray}{0} & 729 \\
    $\textsf{Cachet} + \texttt{sbk05}$ & 13 & 229 & 928 \\
    $\textsf{query-dnnf} + \texttt{bklm16}$ & \textcolor{gray}{0} & 3 & 1017 \\
    \bottomrule
  \end{tabular}
  \label{tbl:tallies}
\end{table}

\begin{figure*}[t]
  \centering
  \input{scatter.tex}%
  \caption{An instance-by-instance comparison between $\textsf{ADDMC} +
    \texttt{cw}$ and the best overall combination of algorithm and encoding
    ($\textsf{Ace} + \texttt{cd06}$, on the left) as well as the second-best
    encoding for \textsf{ADDMC} (\texttt{sbk05}, on the right).}
  \label{fig:scatter}
\end{figure*}

\begin{figure*}[t]
  \centering
  \input{melt.tex}%
  \caption{TODO: describe this in the text}
  \label{fig:melted}
\end{figure*}

In this section, we describe an experimental study comparing the five WMC
encodings for Bayesian networks when run with \textsf{ADDMC}. The experiments
were run on servers with Intel Xeon Gold 6138 and Intel Xeon E5-2630
processors\footnote{Each instance is run on the same processor for all
  encodings.} running Scientific~Linux~7 with a \SI{32}{\gibi\byte} memory limit
and a \SI{1000}{\second} timeout.

@@

@@

\begin{table} % TODO: add bklm16
  \centering
  \caption{Asymptotic upper bounds on the numbers of variables and clauses/ADDs
    for each encoding.}
  \label{tbl:asymptotes}
  \begin{tabular}{lcc}
    \toprule
    Encoding(s) & Variables & Clauses/ADDs \\
    \midrule
    \texttt{cd05}, \texttt{cd06}, \texttt{sbk05} & $O(nv^{d+1})$ & $O(nv^{d+1})$ \\
    \texttt{cw} & $O(nv)$ & $O(nv^2)$ \\
    \texttt{d02} & $O(nv^{d+1})$ & $O(ndv^{d+1})$ \\
    \bottomrule
  \end{tabular}
\end{table}

@@

\bibliography{paper}
